---
title: "Hawaii SNAP Analysis: Interactive Homework"
subtitle: "Using Census Data & Machine Learning to Understand Food Security"
author: "DS 400: Bayesian Statistics & Data Science"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    code-fold: show
    code-tools: true
    df-print: paged
    fig-width: 11
    fig-height: 7
    embed-resources: true
    smooth-scroll: true
execute:
  warning: false
  message: false
  cache: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "100%"
)
```

## 🌺 Introduction: Food Security in Hawaiʻi

Food security is a critical health and equity issue. The Supplemental Nutrition Assistance Program (SNAP) helps vulnerable households afford nutritious food. But which communities have the highest rates of food assistance participation? And what factors are most strongly associated with SNAP participation?

In this assignment, you'll:

1.  **Access Census data** using `tidycensus`
2.  **Explore patterns** of SNAP participation across Hawaiʻi census tracts
3.  **Build a predictive model** using Random Forests
4.  **Interpret results** using SHAP (Explainable AI) techniques

::: {.callout-important icon="📋"}
## Homework Requirements

-   ✅ Complete all **CHALLENGE** sections
-   ✅ Provide code AND written explanations
-   ✅ Show your thinking - comment your code!
-   ✅ Answer reflection questions at the end
-   ⭐ Bonus: Try the extension questions for extra credit

**Due Date**: \[Before Class, 2:30 PM pm 10/23/25\]
:::

------------------------------------------------------------------------

## 📦 Load Libraries

```{r}
#| label: load-packages
#| code-fold: false

library(ggplot2)
library(tidycensus)
library(dplyr)
library(tidyr)
library(stringr)
library(mapgl)
library(viridis)
library(randomForest)
library(treeshap)
library(mapview)
library(sf)
```

------------------------------------------------------------------------

## 📊 Part 1: Collect Census Data

### Set Up Your Census Variables

The code below defines the Census Bureau variables we'll use. These come from the American Community Survey (ACS) 5-Year Estimates.

```{r}
#| label: define-variables
#| code-fold: false

# Define ACS variables for Hawaii
# (Population, poverty, insurance, employment, housing, broadband, education, income, SNAP)

hi_vars <- c(
  # Person-level
  poverty_n            = "B17001_002",
  uninsured_u65_n      = "B27010_017",
  unemployed_n         = "B23025_005",
  pop_total            = "B01003_001",
  
  # Households + housing
  households_total     = "B11001_001",
  broadband_hh_n       = "B28002_004",
  renter_hh_n          = "B25003_003",
  novehicle_hh_n       = "B25044_003",
  snap_hh_n            = "B22001_002",     # <-- SNAP households
  
  # Education
  edu_25plus_total     = "B15003_001",
  bachelors_n          = "B15003_022",
  
  # Labor force + workers
  labor_force_total    = "B23025_003",
  workers_total        = "B08301_001",
  
  # Income level
  median_income        = "B19013_001"
)
```

::: {.callout-note icon="ℹ️"}
## Understanding Census Variables

Each variable code (like "B17001_002") refers to a specific data table and line number from the Census Bureau. The naming convention is: - **B** = Detailed table - **17001** = Table ID (in this case, poverty data) - **002** = Specific line number (usually the numerator for a percentage calculation)

For more info, check the [Census Bureau Data API documentation](https://api.census.gov/data/2023/acs/acs5/variables.html)
:::

### Pull Data from the Census API

```{r}
#| label: get-census-data
#| code-fold: false
#| output: false

# Pull core variables (long format, then pivot to wide)
hi_acs_long <- get_acs(
  geography = "tract",
  state = "HI",
  year = 2023,
  geometry = TRUE,
  variables = hi_vars
)|>
  dplyr::filter(GEOID != "15003981200")

# Convert to wide format (one row per census tract)
hi_acs <- hi_acs_long %>%
  select(GEOID, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate)
```

::: {.callout-note icon="✅"}
Census data loaded successfully! (Output suppressed to keep document clean)
:::

### Calculate Commute Times (≥30 minutes)

```{r}
#| label: commute-data
#| code-fold: false

# Get commute time data using table lookup
commute_tbl <- get_acs(
  geography = "tract",
  state = "HI",
  year = 2023,
  table = "B08303",
  geometry = FALSE
)|>
  dplyr::filter(GEOID != "15003981200")

# Extract commute ≥30 min variable
commute_30plus <- commute_tbl %>%
  filter(variable == "B08303_004") %>%
  group_by(GEOID) %>%
  summarize(commute_30plus_n = sum(estimate, na.rm = TRUE), .groups = "drop")
```

------------------------------------------------------------------------

## 🧮 Part 2: Calculate Percentages & Clean Data

The code below joins all the data and calculates percentages. Study how we normalize each variable!

```{r}
#| label: calculate-percentages
#| code-fold: false

# Join commute data and calculate all percentages
hi_acs <- hi_acs %>%
  left_join(commute_30plus, by = "GEOID") %>%
  mutate(
    # Person-based percentages
    poverty_rate_pct        = 100 * poverty_n       / pmax(pop_total, 1),
    uninsured_u65_pct       = 100 * uninsured_u65_n / pmax(pop_total, 1),
    
    # Labor-force based
    unemployment_rate_pct   = 100 * unemployed_n    / pmax(labor_force_total, 1),
    
    # Household-based percentages
    broadband_pct           = 100 * broadband_hh_n  / pmax(households_total, 1),
    renter_pct              = 100 * renter_hh_n     / pmax(households_total, 1),
    no_vehicle_pct          = 100 * novehicle_hh_n  / pmax(households_total, 1),
    snap_pct                = 100 * snap_hh_n       / pmax(households_total, 1),
    
    # Education (25+)
    bachelors_or_higher_pct = 100 * bachelors_n     / pmax(edu_25plus_total, 1),
    
    # Commute-based
    commute_30plus_pct      = 100 * commute_30plus_n / pmax(workers_total, 1)
  )
```

::: {.callout-tip icon="💡"}
## Why `pmax(denominator, 1)`?

This prevents division by zero errors. If a census tract has 0 people, we use 1 as the denominator instead. The `pmax()` function takes the parallel maximum - comparing each value to 1 element-wise.
:::

------------------------------------------------------------------------

## 🗺️ CHALLENGE 1: Map SNAP Participation

::: {.callout-warning icon="🎯"}
## Your Task

Create a map showing SNAP participation rates by census tract across Hawaiʻi. You should:

1.  **Use `ggplot2`** with `geom_sf()` to map the data
2.  **Color by `snap_pct`** (use a viridis color scale)
3.  **Add labels** including title, subtitle, and data source
4.  **Make it look professional** with clean theme/formatting

:::

```{r}
#| label: challenge-1-snap-map
#| code-fold: false

library(ggplot2)
library(viridis)
library(sf)

ggplot(data = hi_acs) +
  geom_sf(aes(fill = snap_pct), color = "white", size = 0.1) +
  scale_fill_viridis(
    name = "SNAP Participation (%)",
    option = "C",
    direction = -1,
    na.value = "gray90"
  ) +
  labs(
    title = "SNAP Participation Across Hawaiʻi by Census Tract",
    subtitle = "Share of households receiving Supplemental Nutrition Assistance Program (SNAP) benefits, 2023",
    caption = "Source: U.S. Census Bureau, 2023 ACS 5-Year Estimates"
  ) +
  theme_minimal(base_family = "Helvetica") +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray30"),
    plot.caption = element_text(size = 9, color = "gray40", hjust = 0),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  )


```

::: {.callout-note collapse="true"}
## 💭 Reflection on Your Map

After creating your map, answer these questions in a comment:

1.  Where are the highest concentrations of SNAP participation? Big Island, especially in the southern and eastern parts as well as urban areas of Oʻahu, specifically Honolulu’s urban and western neighborhoods.
These areas are shaded in deep purple, indicating higher percentages of households receiving SNAP benefits.
2.  Are there geographic patterns (islands, urban vs. rural)? Urban except for the island of Moloka'i.
3.  What does this tell you about food security in Hawaiʻi? The map shows that food insecurity in Hawaii is influenced by both poverty and location, with the greatest need in rural, isolated areas and in urban communities facing economic hardship.
:::

------------------------------------------------------------------------

## 🔍 CHALLENGE 2: Interactive Map with mapview

::: {.callout-warning icon="🎯"}
## Your Task

Create an **interactive leaflet-style map** using the `mapview` package.

**Hints & Resources:** - The `mapview()` function creates interactive maps - Use `zcol = "snap_pct"` to color by SNAP participation - Set `legend = TRUE` to show a legend - Use `layer.name` to label your layer - The `at` parameter controls color breaks (try: `seq(0, max(...), by = 5)`) - Use `col.regions = viridis::viridis(10, direction = -1)` for colors
:::

```{r}
#| label: challenge-1-snap-map-interactive
#| code-fold: false

library(mapview)
library(viridis)

mapview(
  hi_acs,
  zcol = "snap_pct",                                   # Color by SNAP percentage
  legend = TRUE,                                       # Show legend
  layer.name = "SNAP Participation (%)",               # Legend title
  at = seq(0, max(hi_acs$snap_pct, na.rm = TRUE), 5),  # Color breaks every 5%
  col.regions = viridis(10, direction = -1),           # Viridis color palette
  map.types = "CartoDB.Positron"                       # Clean basemap
)

```

::: {.callout-tip icon="🗺️"}
## Interactive Map Features

With `mapview`, you can: - Zoom and pan - Click on tracts to see exact values - Toggle layers on/off - Switch between map types - Export the map

Try zooming into a specific island and exploring the variation!
:::

------------------------------------------------------------------------

## 📊 CHALLENGE 3: Histogram of SNAP Participation

::: {.callout-warning icon="🎯"}
## Your Task

Create a histogram showing the distribution of `snap_pct` across all Hawaii census tracts.

**Hints & Resources:** - Use `ggplot()` with `geom_histogram()` - Consider bin width - try `bins = 20` or `binwidth = 2` - Add meaningful labels - What does the shape tell you? Is it normally distributed?
:::

```{r}
#| label: challenge-3-snap-histogram
#| code-fold: false

library(ggplot2)

ggplot(hi_acs, aes(x = snap_pct)) +
  geom_histogram(
    bins = 20,
    fill = viridis::viridis(1, option = "C", direction = -1),
    color = "white"
  ) +
  labs(
    title = "Distribution of SNAP Participation Across Hawaiʻi Census Tracts",
    subtitle = "Showing variation in the percentage of households receiving SNAP benefits (2023)",
    x = "SNAP Participation (%)",
    y = "Number of Census Tracts",
    caption = "Source: U.S. Census Bureau, 2023 ACS 5-Year Estimates"
  ) +
  theme_minimal(base_family = "Helvetica") +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray30"),
    plot.caption = element_text(size = 9, color = "gray40", hjust = 0),
    axis.title = element_text(size = 10)
  )




```

::: {.callout-note collapse="true"}
## 📈 Interpretation Questions

1.  Is the distribution symmetric or skewed? The distribution is strongly right-skewed
2.  What's the typical SNAP participation rate? Most tracts fall between 0% and 15%, suggesting that typical SNAP participation is relatively low across much of Hawaiʻi.
3.  Are there any outliers or surprising patterns? Yes, there are a few tracts with very high participation rates (above 75, 100%), which stand out from the rest.
:::

------------------------------------------------------------------------

## 🌲 Part 3: Machine Learning Setup

Now we'll prepare for predictive modeling. First, create the ML-ready dataset:

```{r}
hi_acs_data_ml <- hi_acs %>%
  st_drop_geometry() %>%                          
  select(
    median_income, poverty_rate_pct, uninsured_u65_pct, 
    unemployment_rate_pct, broadband_pct, renter_pct, no_vehicle_pct, 
    snap_pct, bachelors_or_higher_pct, commute_30plus_pct
  ) %>%
  drop_na()                                       

```

------------------------------------------------------------------------

## 🎯 CHALLENGE 4: Build a Random Forest Model

::: {.callout-warning icon="🎯"}
## Your Task

Train a Random Forest to predict SNAP participation rates using the other socioeconomic indicators.

**Your Model Should:** - Use `snap_pct` as the target variable (left side of `~`) - Use all other variables as predictors (`.` on right side) - Have `ntree = 500` for stability - Be named `rf`

**Hints & Resources:** - Use the `randomForest()` function - Formula syntax: `target ~ .` predicts target using all other columns - Structure: `randomForest(formula, data = ..., ntree = 500)` - Print the model to see performance metrics!
:::

```{r}
#| label: challenge-4-random-forest
#| code-fold: false

library(randomForest)

rf <- randomForest(
  snap_pct ~ .,                  
  data = hi_acs_data_ml,
  ntree = 500,                   
  importance = TRUE              
)

print(rf)


```

::: {.callout-note collapse="true"}
## 🔍 Understanding Your Model Output

Look at the printed output and answer:

1.  **Mean of squared residuals (MSR)**: How accurate are the predictions on average? The model’s predictions differ from the actual SNAP participation rates by about √45.83 ≈ 6.8 percentage points. That’s a moderate level of prediction error.
2.  **% Var explained**: What percentage of SNAP variation does the model explain? The model explains 58.6% of the variation in SNAP participation across census tracts.
3.  **Interpretation**: Is this good? What might the model be missing? This is a fairly strong model for social and economic data, where perfect prediction is rare. However, it may miss important factors not included in the Census data, such as food prices, store access, local programs, or cultural influences.

Remember: MSR tells you the average squared prediction error. % Var explained is like R² - higher is better!
:::

------------------------------------------------------------------------

## 🔬 Part 4: Explainable AI with SHAP

SHAP (SHapley Additive exPlanations) values explain *why* the model makes specific predictions. They break down each prediction into individual contributions from each feature.

### Understanding SHAP

Before we calculate, let's understand what we're doing:

1.  **Unify the model**: Convert the Random Forest into a format that SHAP can work with
2.  **Calculate SHAP values**: Compute explanations for a subset of observations (computing for all can be slow)
3.  **Interpret**: Use the results to explain individual predictions and global patterns

------------------------------------------------------------------------

## 🎯 CHALLENGE 8: Prepare SHAP Calculations

::: {.callout-warning icon="🎯"}
## Your Task

Prepare the model for SHAP analysis by:

1.  **Unifying** your Random Forest model using `unify()`
2.  **Calculating SHAP values** for the first 200 observations using `treeshap()`

**Hints & Resources:**

-   Step 1: Use `unify(your_model, your_data)` to convert the RF
    -   First argument: your Random Forest object (named `rf`)
    -   Second argument: your ML dataset (named `hi_acs_data_ml`)
    -   Assign result to `unified`
-   Step 2: Use `treeshap(unified_object, data_subset, verbose = 0)`
    -   First argument: your unified model
    -   Second argument: subset of data - use `hi_acs_data_ml[1:200, ]` to get first 200 rows
    -   Third argument: `verbose = 0` suppresses progress messages
    -   Assign result to `treeshap1`
:::

```{r}
unified <- unify(rf, hi_acs_data_ml)
treeshap1 <- treeshap(unified, hi_acs_data_ml[1:200, ], verbose = 0)
treeshap1

```

::: {.callout-note collapse="true"}
## 💭 Why Only 200 Observations?

Computing SHAP values is computationally expensive - it can take a long time for large datasets. For this assignment, we're calculating for a representative sample of 200 census tracts, which is enough to: - Understand which features matter globally - Examine feature dependence patterns - Look at individual case examples

In production, you might calculate for all observations or use approximation methods like TreeSHAP (which is what we're using!).
:::

::: {.callout-tip icon="⏱️"}
## ⚠️ Patience Required!

The SHAP calculation might take 30 seconds to a few minutes depending on your computer. The `verbose = 0` parameter keeps it quiet while it works. You'll know it's done when the next code chunk starts executing!
:::

------------------------------------------------------------------------

## 📊 CHALLENGE 5: Feature Importance Plot

::: {.callout-warning icon="🎯"}
## Your Task

Create a feature importance plot showing the top 5 most important variables for predicting SNAP participation.

**Hints & Resources:** - Use `plot_feature_importance(treeshap1, max_vars = 5)` - This uses the SHAP values we just calculated - The plot shows which features have the biggest impact on predictions - Longer bars = more important
:::

```{r}
library(treeshap)

plot_feature_importance(treeshap1, max_vars = 5)


```

::: {.callout-note collapse="true"}
## 💭 Interpretation

Based on your feature importance plot:

1.  What are the TOP 2 most important predictors of SNAP participation? The top two predictors of SNAP participation are the percentage of residents with a bachelor’s degree or higher and the poverty rate.
2.  Why do you think these variables matter? These variables matter because they both reflect economic opportunity and need. Areas with fewer college-educated residents often have lower incomes and higher poverty, increasing reliance on SNAP benefits.
3.  Are there any surprising results? It is surprising that education level is even more influential than poverty rate
4.  What does this tell you about food security in Hawaiʻi? This suggests that food security in Hawaiʻi is deeply connected to both economic inequality and access to education.
:::

------------------------------------------------------------------------

## 📈 CHALLENGE 6: Feature Dependence Plots

::: {.callout-warning icon="🎯"}
## Your Task

Create feature dependence plots for the **top 5 most important features** you identified in Challenge 5.

**What to Do:** 1. Look at your feature importance plot 2. Identify the 5 most important variables 3. Create a dependence plot for each one 4. Use: `plot_feature_dependence(treeshap1, "variable_name")`

**Hints & Resources:** - The x-axis shows the actual variable values - The y-axis shows SHAP values (how much it pushes predictions up/down) - Red dots = push toward higher SNAP % - Blue dots = push toward lower SNAP % - Look for patterns: linear? non-linear? threshold effects?

**Your Code (plot all 5):**
:::

```{r}
#| label: challenge-6-dependence-1
#| code-fold: false

plot_feature_dependence(treeshap1, "poverty_rate_pct")

```

```{r}
#| label: challenge-6-dependence-2
#| code-fold: false

plot_feature_dependence(treeshap1, "median_income")

```

```{r}
#| label: challenge-6-dependence-3
#| code-fold: false

plot_feature_dependence(treeshap1, "unemployment_rate_pct")

```

```{r}
#| label: challenge-6-dependence-4
#| code-fold: false

plot_feature_dependence(treeshap1, "bachelors_or_higher_pct")

```

```{r}
#| label: challenge-6-dependence-5
#| code-fold: false

plot_feature_dependence(treeshap1, "broadband_pct")

```

::: {.callout-note collapse="true"}
## 📊 Dependence Analysis Questions

For each plot, write 1-2 sentences about:

1.  **Direction**: Does higher variable value → higher or lower SNAP participation?
2.  **Strength**: Is the relationship strong, moderate, or weak?
3.  **Pattern**: Is it linear, or are there non-linear patterns?
4.  **Outliers**: Are there interesting outlier tracts?

Example answer: \> "Poverty rate shows a strong positive relationship with SNAP participation. As poverty increases, SHAP values increase substantially, pushing predictions toward higher SNAP participation. The relationship appears roughly linear."
:::

------------------------------------------------------------------------

## 👁️ CHALLENGE 7: Individual Tract Contribution

::: {.callout-warning icon="🎯"}
## Your Task

Examine what drives predictions for **Census Tract 304.04 on Maui, Hawaiʻi**.

**Hints & Resources:** - This tract is observation #154 in our data (hint: `obs = 154`) - Use: `plot_contribution()` - **Red bars** = pushing toward higher SNAP % - **Blue bars** = pushing toward lower SNAP % - **Bar length** = strength of effect

**Your Code:**
:::

```{r}
#| label: challenge-7-contribution
#| code-fold: false

# Plot contribution effects for Census Tract 304.04 (Maui, observation #154)
plot_contribution(treeshap1, obs = 154)


```

::: {.callout-note collapse="true"}
## 🔍 Case Study Analysis

Answer these questions about Maui Census Tract 304.04:

1.  **Baseline**: What's the model's baseline prediction? The model’s baseline represents the average predicted SNAP participation across all tracts in Hawaii.
2.  **Pushers**: Which 2-3 variables push predictions toward HIGHER SNAP%? Variables such as high poverty rate, low median income, and lower broadband access likely push the prediction toward a higher SNAP percentage.
3.  **Pullers**: Which 2-3 variables push predictions toward LOWER SNAP%? Features like a higher education level (bachelor’s or higher) and greater broadband access can pull the prediction toward a lower SNAP percentage.
4.  **Net Effect**: Is the final prediction above or below the baseline? The final prediction is above the baseline which means that the model expects this area to have higher than average SNAP participation.
5.  **Story**: Write 2-3 sentences telling the "story" of this tract's SNAP participation. Economic challenges such as poverty and limited income likely drive greater reliance on SNAP benefits. Some stabilizing factors like education or connectivity may help reduce need, showing a community balancing both opportunity and hardship.
:::

------------------------------------------------------------------------

## 🎓 Reflection & Analysis

::: {.callout-important icon="🤔"}
## Final Reflection Questions

Answer these questions in text below:

### Understanding the Analysis

1.  **Model Performance**: How well does your Random Forest predict SNAP participation? What might explain the remaining unexplained variation? The model predicted SNAP participation fairly well. It captured most of the major patterns (~59%) but not everything. The remaining variation could be due to factors not included in the dataset.

2.  **Key Drivers**: Based on feature importance and dependence plots, which 2-3 factors are most strongly associated with SNAP participation in Hawaiʻi? The most important predictors of SNAP participation were poverty rate, median income, and unemployment rate. These make sense because lower income and higher unemployment usually lead to more people needing food assistance.

3.  **Equity Implications**:

    -   What does the geographic distribution of SNAP participation tell you about food security disparities?
    -   Are certain communities more vulnerable? The map showed that SNAP participation is higher in both rural and urban areas under economic pressure. This suggests that food insecurity in Hawaiʻi is shaped by both poverty and geography.

4.  **Limitations**:

    -   What important variables might be missing from this analysis?
    -   What are the limitations of predicting SNAP participation? (Remember: correlation ≠ causation) This analysis doesn’t include every factor that affects food insecurity, such as food prices, local store access, or community support programs. Also, the model only shows correlations, not direct causes.

5.  **Policy Recommendations**:

    -   Based on your findings, what 2-3 policy interventions might help address food insecurity? Improving job access, supporting local food systems, and expanding outreach in high need areas.
    -   Who should be targeted? Where should resources be focused? Efforts should target both low income urban neighborhoods and remote rural communities where access and opportunity are limited.

### Technical Reflection

6.  **Machine Learning Interpretation**: How did SHAP values help you understand the Random Forest model compared to just looking at variable importance? They helped me see why the model made certain predictions. Instead of just knowing which variables were important, it showed me how each feature pushed the prediction higher or lower.

7.  **Next Steps**: If you were to continue this analysis, what would you do next? I’d gather more detailed community data, like information on local food programs, transportation access, and housing costs.
:::

------------------------------------------------------------------------

## ⭐ BONUS: Extension Challenges

::: {.callout-tip icon="🚀"}
## For Extra Credit

Try one or more of these extensions:

### Extension 1: Regional Analysis

-   Split Hawaiʻi by island (using `GEOID` or geographic location)
-   Build separate Random Forest models for each island
-   Compare which factors matter most on each island

### Extension 2: Relationship Exploration

-   Create scatter plots showing relationships between top predictors and SNAP%
-   Calculate correlation coefficients
-   Which relationships are strongest?

### Extension 3: Model Comparison

-   Build a **second** Random Forest with different parameters (e.g., `mtry`, `nodesize`)
-   Compare performance to your original model
-   Which performs better?

### Extension 4: Prediction Intervals

-   For a new hypothetical census tract, use your model to predict SNAP participation
-   What would be needed to create prediction intervals?
-   How confident would you be in that prediction?

### Extension 5: Data Visualization

-   Create a multi-panel visualization combining:
    -   The map of SNAP participation
    -   Distribution histogram
    -   Top feature importance plot
    -   Tell a complete visual story
:::

------------------------------------------------------------------------

## 📝 Submission Checklist

-   [ ] All 8 challenges completed with code
-   [ ] Written explanations for each challenge
-   [ ] Reflection questions answered thoroughly
-   [ ] Code is commented and readable
-   [ ] Plots are properly labeled with titles/legends
-   [ ] At least one bonus challenge attempted (optional)
-   [ ] **Rendered to PDF and submitted to Canvas**

------------------------------------------------------------------------

## 📤 How to Render & Submit

### Step 1: Render to HTML

In RStudio, click the blue **"Render"** button at the top of the editor pane.

Wait for the rendering to complete - you'll see the document appear in the **Viewer pane** (bottom right).

::: {.callout-tip icon="⏱️"}
**Rendering typically takes 2-5 minutes depending on how fast your computer is. Be patient!**
:::

### Step 2: Open in New Window

In the **Viewer pane**, click the **"Show in new window"** icon (it looks like an arrow pointing to a box, usually in the top-right corner of the viewer).

Your rendered HTML document will open in your default web browser.

### Step 3: Print to PDF

1.  In your browser, press **`Ctrl + P`** (Windows) or **`Cmd + P`** (Mac)
2.  A print dialog will appear
3.  Change the destination to **"Save as PDF"**
4.  Give it a clear name: `LastName_FirstName_ml_assignment.pdf`
5.  Click **"Save"**

Your PDF will download to your Downloads folder (or wherever your browser saves files).

### Step 4: Upload to Canvas

1.  Go to your Canvas course page
2.  Find the assignment: "Hawaii SNAP Analysis Homework"
3.  Click **"Submit Assignment"**
4.  Click **"Choose File"** and select your PDF
5.  Click **"Submit"**

::: {.callout-important icon="⚠️"}
## Important Reminders

-   ✅ Make sure your `.qmd` file renders **without errors** before saving as PDF
-   ✅ Double-check that all your challenge code is visible in the HTML output
-   ✅ Your PDF should include both your code AND the results/plots
-   ✅ Save the PDF file somewhere you can find it (Desktop or a course folder)
-   ⏱️ Allow 5-10 minutes total for rendering and saving
:::

### Troubleshooting

**"Render button is grayed out?"** - Make sure you have the `.qmd` file open in the editor (not just viewed)

**"I don't see a Viewer pane?"** - Go to `View` → `Panes` → `Show Viewer Pane`

**"Nothing appears after clicking Render?"** - Check the R console for error messages - Make sure all your code runs without errors first - Try re-rendering after fixing any issues

------------------------------------------------------------------------

## 📚 Resources & References

-   [tidycensus Documentation](https://walker-data.com/tidycensus/)
-   [Census Bureau Data Catalog](https://data.census.gov/)
-   [SHAP Documentation](https://shap.readthedocs.io/)
-   [Random Forests in R](https://cran.r-project.org/web/packages/randomForest/randomForest.pdf)
-   [mapview Package](https://r-spatial.github.io/mapview/)

------------------------------------------------------------------------

*Good luck! Remember: focus on understanding WHAT you're doing and WHY, not just running code. Ask questions, explore the data, and enjoy the detective work of data science! 🔍📊*
